Expert-In-The-Loop Causal Discovery: Iterative Model Refinement Using Expert Knowledge

### 2025-02-05 paper/main.tex
Updates

In the previous section, we discussed how different types of variables require
different conditional measures of association. However, no unified measure
exists for mixed data. In this section, we introduce a novel measure of partial
association for mixed data, by extending the idea of partial correlation
coefficient to mixed data. Specifically, our approach combines a mixed data
residualization method \citep{Ankan2023} with Pillai's Trace
\citep{Pillai1955}, a multivariate measure of association based on canonical
correlations.

@@

Given $ n $ samples $ D = (x, y, \bm{z}) $ of variables $ X $, $ Y $, and $
\bm{Z} $, we aim to estimate the conditional association $ \rho(X, Y; \bm{Z}) $
between $ X $ and $ Y $ conditioned on $ \bm{Z} $. We first use the
residualization approach from \citet{Ankan2023} to compute the residual
matrices for both $ X $ and $ Y $. For each of the variable $ X $ and $ Y $, we
compute the residual $ R_X $ and $R_Y$ respectively as follows depending on
their type.

\begin{enumerate}
	\item \textbf{Continuous:} We train a regression model, $ E_X: x \sim
		\bm{z} $. The residuals are then defined as the difference
		between the true and the predicted values using $ E_X $. 
		$$ R_{x_i} = x_i - E_X(\bm{z}_i) $$
	\item \textbf{Ordinal:} We start by training a probability estimator, $
		p_X: x \sim \bm{z} $, and obtain probability predictions 
		$ \hat{p}_X: \hat{p}_X(\bm{Z}) $. Then, we compute the residuals
		as follows \citep{Li2012}:
		$$ R_{x_i} = \hat{p}(\hat{x}_i > x_i) - \hat{p}(\hat{x}_i < x_i) $$
	\item \textbf{Categorical:} We again start by training a probability
		estimator $ p_X: X \sim \bm{Z} $, and obtain probability
		estimates $ \hat{p}_X: p_Z(\bm{Z}) $. Next, we dummy encode the
		categorical variable and then compute the residuals as follows: 
		$$ R_X = x_i - \hat{p}(\bm{z}_i) $$
\end{enumerate}

\todo[inline]{Fix notations above}

We repeat the above steps for both the variables $ X $ and  $ Y $ to get
residuals $ R_X $ and $ R_Y $. Here, we have the option to choose the
estimators based on the properties of the data. Non-parametric ensemble
estimators such as Random Forest and XGBoost can handle any combination of data
types while performing reasonably well. For linear datasets, we can also choose
variants of linear regression depending on the type of data.

After the above steps are repeated for both $ X $ and $ Y $, we end up with two
residual matrices: $ R_X $ and $ R_Y $. The type of the variable determines the
shape for these matrices. If the variable is continuous, we get a residual
matrix of shape $ ( n \times 1 ) $ and if the variable is categorical, we have
a residual matrix of shape $ ( x \times (k -1)) $ where one of the dummy encoded
variables is dropped to avoid multicollinearity.

We now have two continuous-valued residual matrices. The next step is
to quantify the association between these residuals. In multivariate
statistics, canonical correlation based measures have been used to measure
the association between two sets of variables. We can treat our residuals
as sets of variables and use canonical correlations to quantify the association
between them.

@@

	% \begin{equation}
	% 	\phi(\bm{U}, \bm{V}) = \sup_{a, b} \frac{a^T \Sigma_{\bm{U}\bm{V}} b}{\sqrt{a^T \Sigma_{\bm{U}\bm{U}} a} \sqrt{b^T \Sigma_{\bm{V}\bm{V}} b}}
	% \end{equation}

% \begin{equation}
% 	\begin{split}
% 		\bm{\rho}(\bm{U}, \bm{V}) &= \sqrt{\mathrm{E} \left( \Sigma_{\bm{UU}}^{-1} \Sigma_{\bm{UV}} \Sigma_{\bm{VV}}^{-1} \Sigma_{\bm{VU}} \right)}
% 	\end{split}
% \end{equation}
	\begin{equation}
		\rho(\bm{U}, \bm{V}) = \max_{a, b} Corr(a^T \bm{U}, b^T \bm{V})
	\end{equation}

@@

% where $ \mathrm{E}(\Sigma) $ denotes the eigenvalues of the matrix $ \Sigma $.
% $ \Sigma_{UU} $ and $ \Sigma_{VV} $ are the covariance matrices of $ \bm{U} $
% and $ \bm{V} $, and $ \Sigma_{UV} $ and $ \Sigma_{VU} $ are the
% cross-covariance matrices.

Canonical correlations generalize correlations to multi-dimensional variables. They
find orthogonal linear transformations $ a $ and $ b $ of $ \bm{U} $ and $ \bm{V} $
such that the correlation between the transformed variables $ a^T \bm{U} $ and 
$ b^T \bm{V} $ is maximized. This yields a vector of correlation values of 
size $ \min(p, q) $, representing the correlation coefficient of each pair of 
transformed variables. Pearson's correlation coefficient is a special case of 
canonical correlations when $ p = q = 1 $.

@@

This measure of association has many desired properties:

>>>>

In the previous section, we discussed how different types of variables require
different conditional measures of association. However, no unified measure
exists for mixed data. In this section, we introduce a novel measure of
conditional association for mixed data by extending the idea of partial
correlation coefficient to mixed data. Specifically, our approach combines a
mixed data residualization method \citep{Ankan2023} with Pillai's Trace
\citep{Pillai1955}, a multivariate measure of association based on canonical
correlations.

@@

Given a dataset $ D = (x, y, \bm{z}) $ on variables $ X $, $ Y $, and $ \bm{Z}
$, we aim to estimate the conditional association $ \rho(X, Y; \bm{Z}) $
between $ X $ and $ Y $ conditioned on $ \bm{Z} $. We first use a mixed data
residualization approach \citep{Ankan2023} to compute residuals $R_X$ and $R_Y$
for $ X $ and $ Y $. The method for computing these residuals depend on the 
type of variable:

\begin{enumerate}
	\item \textbf{Continuous:} We train a model, $ E_X: x \sim
		\bm{z} $. The residuals are then computed by taking the difference
		between the true and the predicted values using $ E_X $. 
		$$ R_{x_i} = x_i - E_X(\bm{z}_i) $$
	\item \textbf{Ordinal:} We start by training a probability estimator, $
		p_X: x \sim \bm{z} $, and then use the estimated probabilities, 
		$ \hat{p}_X(x) $ to compute the residuals:
		$$ R_{x_i} = \hat{p}_X(X < x_i) - \hat{p}_X(X > x_i) $$
	\item \textbf{Categorical:} We again start by training a probability
		estimator $ p_X: X \sim \bm{Z} $, and obtain probability
		estimates $ \hat{p}_X: p_X(\bm{Z}) $. Next, we dummy encode the
		categorical variable and then compute the residuals as follows: 
		$$ R_X = x_i - \hat{p}(\bm{z}_i) $$
\end{enumerate}

We repeat the above steps for both the variables $ X $ and  $ Y $ to get
residuals $ R_X $ and $ R_Y $. Here, we have the option to choose the
estimators based on the properties (distribution, type of relationship, and so
on) of the data. Non-parametric ensemble estimators such as Random Forest and
XGBoost can handle any combination of data types while performing reasonably
well on a wide range of distributions and relationship types. For linear
relationships, we can also choose simpler models such as linear regression.

After the above residualization steps are repeated for both $ X $ and $ Y $, we
end up with two residual matrices: $ R_X $ and $ R_Y $. The type of the
variable determines the shape for these matrices. If the variable is continuous
or ordinal, we get a residual matrix of shape $ ( n \times 1 ) $ and if the
variable is categorical, we have a residual matrix of shape $ ( x \times (k-1))
$ where one of the dummy encoded variables is dropped to avoid
multicollinearity.

We now have two continuous-valued residual matrices. The next step is
to quantify the association between these residuals. For this purpose
we use canonical correlations that have been used to measure the association
between two sets of random variables.

@@

	\begin{equation}
		\rho(\bm{U}, \bm{V}) = \max_{a, b} Corr(a^T \bm{U}, b^T \bm{V})
	\end{equation}

@@

Canonical correlations generalize correlation coefficients to multi-dimensional
variables. They find orthogonal linear transformations $ a $ and $ b $ of $
\bm{U} $ and $ \bm{V} $ such that the correlation coefficient between the
transformed variables $ a^T \bm{U} $ and $ b^T \bm{V} $ is maximized. This
yields a vector of correlation coefficient values of size $ \min(p, q) $,
representing the correlation coefficient of each pair of transformed variables.
Pearson's correlation coefficient is a special case of canonical correlations
when $ p = q = 1 $.

@@

This measure of association has many desired properties for our application:
